---
title: "Air Quality Dashboard - EPA Annual Concentrations 2024"
author: "Yusi Yao, Xavier Akers, Keshavd"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    self_contained: true
    theme:
      version: 5
      bootswatch: flatly
      base_font:
        google: Prompt
      code_font:
        google: JetBrains Mono
---

```{r setup, include=FALSE}
library(flexdashboard)
# Install thematic and un-comment for themed static plots (i.e., ggplot2)
# thematic::thematic_rmd()

```

```{r, echo=FALSE, message=FALSE}
library(readr)
library(ggplot2)
library(leaflet)
library(dplyr)
library(kableExtra)
library(DT)
data <- read_csv("data/annual_conc_by_monitor_2024.csv")

```

# Introduction & Abstract


Column {data-width=350}
-----------------------------------------------------------------------

### Abstract

This dashboard analyzes the EPA Annual Concentration dataset for 2024, summarizing pollutant measurements across U.S. monitoring sites. We explore geographic patterns, pollutant trends, and apply statistical models to understand air quality outcomes and predictors.

### Data Dictionary

```{r}
variables <- colnames(data)

descriptions <- c(
  "State FIPS code", "County FIPS code", "Monitoring site number", "Parameter code for pollutant",
  "Point of contact", "Latitude coordinate", "Longitude coordinate", "Geodetic datum",
  "Pollutant name", "Sample duration", "Pollutant standard reference", "Metric used for analysis",
  "Measurement method", "Year of data", "Units of measure", "Event type recorded",
  "Number of valid observations", "Percent of valid observations", "Data completeness flag",
  "Number of valid days", "Number of required days", "Exceptional data count",
  "Null data count", "Primary standard exceedances", "Secondary standard exceedances",
  "Certification status", "Number of observations below detection limit",
  "Arithmetic mean concentration", "Arithmetic standard deviation", "Maximum observed value",
  "Datetime of max value", "Second max value", "Datetime of second max", "Third max value",
  "Datetime of third max", "Fourth max value", "Datetime of fourth max", 
  "First non-overlapping max", "Datetime of first non-overlapping max", 
  "Second non-overlapping max", "Datetime of second non-overlapping max", 
  "99th percentile", "98th percentile", "95th percentile", "90th percentile", 
  "75th percentile", "50th percentile", "10th percentile", "Local site name",
  "Monitoring site address", "State name", "County name", "City name",
  "Core-Based Statistical Area name", "Date of last data change"
)

data_dictionary <- data.frame(
  Variable = variables,
  Description = descriptions[1:length(variables)]
)

kbl(data_dictionary) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Column {data-width=650}
-----------------------------------------------------------------------

### Map of Monitoring Stations

```{r}
leaflet(data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~Longitude, lat = ~Latitude,
    radius = 1,
    popup = ~paste("Pollutant:", `Parameter Name`, "<br>Mean:", `Arithmetic Mean`)
  )
```


# Multiple Regression

Column {data-width=650}
-------------------------------------

### Description

1. Multiple Linear Regression

Research Question:
Can we predict the annual average concentration (Arithmetic Mean) of PM2.5 based on location and temporal variables?

Modeling Approach:
	•	Response Variable: Arithmetic Mean
	•	Predictor Variables: Latitude, Longitude, Year, Observation Count, 1st Max Value, 90th Percentile, 10th Percentile

Justification:
This model aims to understand how geographical and temporal factors influence PM2.5 concentrations.

---








# Ridge Regression

Column {data-width=650}
-------------------------------------
2. Ridge Regression (using glmnet)

Research Question:
Does incorporating multiple correlated predictors improve the prediction of PM2.5 concentrations?

Modeling Approach:
- Response Variable: Arithmetic Mean
- Predictor Variables: 1st Max Value, 2nd Max Value, 3rd Max Value, 4th Max Value, 90th Percentile, 75th Percentile, 50th Percentile, 10th Percentile, Observation Count

Justification:
Ridge regression is suitable for handling multicollinearity among predictors, potentially enhancing model performance.

---

# Natural Spline or LOESS

Column {data-width=650}
-------------------------------------

3. Natural Cubic Spline or LOESS Fit

Research Question:
How do PM2.5 concentrations vary over time within a specific location?

Modeling Approach:
- Response Variable: Arithmetic Mean
- Predictor Variable: Date (converted from 1st Max DateTime)
- Method: Apply a natural cubic spline or LOESS smoothing to model temporal trends.

Justification:
This approach captures non-linear temporal patterns in pollutant concentrations.

---

# kNN

Column {data-width=650}
-------------------------------------

4. k-Nearest Neighbors (kNN) Classification

Research Question:
Can we classify days as ‘High’ or ‘Low’ pollution based on observed metrics?

Modeling Approach:
	•	Response Variable: Pollution Level (categorical: ‘High’ if Arithmetic Mean > threshold, else ‘Low’)
	•	Predictor Variables: 1st Max Value, 90th Percentile, Observation Count, Latitude, Longitude

Justification:
kNN can classify pollution levels based on similarity in observed features.

---

# Naive Bayes

Column {data-width=650}
-------------------------------------
5. Naive Bayes Classification

Research Question:
Can we predict the Parameter Name (type of pollutant) based on statistical features?

Modeling Approach:
	- Response Variable: Parameter Name
	- Predictor Variables: Arithmetic Mean, 1st Max Value, 90th Percentile, Observation Count

Justification:
Naive Bayes is effective for multi-class classification problems with categorical outcomes.

---

# Logistic Regression

Column {data-width=650}
-------------------------------------

6. Logistic Regression Classification

Research Question:
What factors influence the likelihood of exceeding primary pollutant standards?

Modeling Approach:
	- Response Variable: Exceedance (binary: 1 if Primary Exceedance Count > 0, else 0)
	- Predictor Variables: Arithmetic Mean, 1st Max Value, Observation Count, Parameter Name (encoded)

Justification:
Logistic regression models the probability of a binary outcome based on predictor variables.


Page 1 {data-navmenu="Menu A" data-navmenu-icon="fa-bookmark"}
=====================================


Page 2 {data-navmenu="Menu A"}
=====================================  


Page 3 {data-navmenu="Menu B" data-navmenu="fa-cog"}
=====================================


Page 4 {data-navmenu="Menu B"}
=====================================  

Page 1
===================================== 

You can link to a dashboard page with either of the following syntaxes:

[Page 2]

[Page Two](#page-2)

The second syntax is used when you want a custom name for the link 
(rather than just using the page title).

### Chart A
    
```{r}
```

Page 2
=====================================     

### Chart B
    
```{r}
```


