---
title: "U.S. Environment Insight"
author: "Yusi Yao, Xavier Akers, Keshav Dwivedi"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    self_contained: true
    theme:
      version: 5
      bootswatch: lumen
      base_font: '"Open Sans", -apple-system, BlinkMacSystemFont, "Segoe UI", Roboto, Helvetica, Arial, sans-serif'
      code_font:
        google: JetBrains Mono
editor_options: 
  markdown: 
    wrap: 72
---


<style>

.navbar {
    margin-bottom: 5px;  /* default is around 20px */
}

.section {
    margin-top: 5px;  /* tighten vertical spacing */
}

</style>



```{r setup, include=FALSE}
library(flexdashboard)
library(readr)
library(ggplot2)
library(leaflet)
library(dplyr)
library(kableExtra)
library(DT)
library(broom)
library(viridisLite)
library(splines)
library(e1071)
library(caret)
library(glmnet)
library(lubridate)
library(highcharter)

data <- read_csv("data/annual_conc_by_monitor_2024.csv")

# Define custom Highcharter theme based on hc_theme_economist
thm <- hc_theme_merge(
  hc_theme_economist(),
  hc_theme(
    chart = list(
      backgroundColor = "#ffffff",
      style = list(fontFamily = "Open Sans")
    ),
    title = list(style = list(fontFamily = "Open Sans", fontSize = "18px")),
    xAxis = list(
      gridLineWidth = 1,
      labels = list(style = list(fontFamily = "Open Sans"))
    ),
    yAxis = list(
      labels = list(style = list(fontFamily = "Open Sans"))
    ),
    plotOptions = list(
      series = list(
        animation = list(duration = 1000),
        dataLabels = list(style = list(fontFamily = "Open Sans"))
      )
    )
  )
)

# Define a professional color palette
colors <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd")
```







Introduction & Abstract
===================================== 


Column {data-width=350}
-------------------------------------

### Abstract

Welcome to the 2024 U.S. Environment Insights Dashboard. We analyze the 2024 EPA Annual Concentration dataset to reveal important insights about the air quality in the United States. By examining pollutant measurements from monitoring sites aaround the country, we find trends in the data that are significant for public health and environmental policy. Our statistical methods predict the results of air quality and provide insight on the factors that affect them. This data-driven initiative aligns with EPA's mission to protect the environment and achieve cleaner air for the public.

The data can be downloaded here [annual_conc_by_moinitor_2024.csv](https://drive.google.com/file/d/1VA5CdbeluNWSM6bxj-k1c8vU2GCBBbZi/view?usp=sharing) and is originally sourced from [EPA Pre-Generated Data File](https://aqs.epa.gov/aqsweb/airdata/download_files.html).

### Data Dictionary

```{r data_dictionary}
variables <- colnames(data)

descriptions <- c(
  "State FIPS code", "County FIPS code", "Monitoring site number", "Parameter code for pollutant",
  "Point of contact", "Latitude coordinate", "Longitude coordinate", "Geodetic datum",
  "Pollutant name", "Sample duration", "Pollutant standard reference", "Metric used for analysis",
  "Measurement method", "Year of data", "Units of measure", "Event type recorded",
  "Number of valid observations", "Percent of valid observations", "Data completeness flag",
  "Number of valid days", "Number of required days", "Exceptional data count",
  "Null data count", "Primary standard exceedances", "Secondary standard exceedances",
  "Certification status", "Number of observations below detection limit",
  "Arithmetic mean concentration", "Arithmetic standard deviation", "Maximum observed value",
  "Datetime of max value", "Second max value", "Datetime of second max", "Third max value",
  "Datetime of third max", "Fourth max value", "Datetime of fourth max", 
  "First non-overlapping max", "Datetime of first non-overlapping max", 
  "Second non-overlapping max", "Datetime of second non-overlapping max", 
  "99th percentile", "98th percentile", "95th percentile", "90th percentile", 
  "75th percentile", "50th percentile", "10th percentile", "Local site name",
  "Monitoring site address", "State name", "County name", "City name",
  "Core-Based Statistical Area name", "Date of last data change"
)

data_dictionary <- data.frame(
  Variable = variables,
  Description = descriptions[1:length(variables)]
)

kbl(data_dictionary) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Column {data-width=650}
-------------------------------------

### Map of Monitoring Stations

```{r monitoring_map}
# Aggregate data by unique monitoring site
map_data <- data %>%
  mutate(site_id = paste(`State Code`, `County Code`, `Site Num`, sep = "-")) %>%
  group_by(site_id) %>%
  summarise(
    Longitude = mean(Longitude, na.rm = TRUE),
    Latitude = mean(Latitude, na.rm = TRUE),
    Mean_Concentration = mean(`Arithmetic Mean`, na.rm = TRUE),
    Common_Pollutant = names(sort(table(`Parameter Name`), decreasing = TRUE))[1],
    .groups = "drop"
  ) %>%
  filter(!is.na(Longitude) & !is.na(Latitude) & !is.na(Mean_Concentration) &
         Longitude >= -180 & Longitude <= 180 & Latitude >= -90 & Latitude <= 90)

pal <- colorNumeric(
  palette = "YlOrRd",
  domain = map_data$Mean_Concentration
)

leaflet(map_data, options = leafletOptions(minZoom = 3)) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(
    lng = ~Longitude, lat = ~Latitude,
    radius = 3,
    color = ~pal(Mean_Concentration),
    clusterOptions = markerClusterOptions(
      spiderfyOnMaxZoom = TRUE,
      showCoverageOnHover = TRUE,
      zoomToBoundsOnClick = TRUE
    ),
    popup = ~paste("Pollutant:", Common_Pollutant, "<br>Mean:", round(Mean_Concentration, 2))
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~Mean_Concentration,
    title = "Arithmetic Mean (µg/m³)",
    opacity = 1
  )
```
















Multiple Regression {data-navmenu="MLR" data-icon="fa-chart-line"}
===================================== 

Column {data-width=500}
-------------------------------------

### Multiple Linear Regression

**Research Question**: Can we predict the annual average concentration (Arithmetic Mean) of PM2.5 based on location and temporal variables?

**Modeling Approach**:  
- **Response Variable**: Arithmetic Mean  
- **Predictor Variables**: Latitude, Longitude, Year, Observation Count, 1st Max Value, 90th Percentile, 10th Percentile  

**Justification**: This model aims to understand how geographical and temporal factors influence PM2.5 concentrations.

**Discussion**: Our multiple linear regression investigate whether annual average PM2.5 (Arithmetic Mean) can be predicted from coordinates, observation count, and pollutant measurement percentiles.

We find that all predictors except Year show statistically significant relationships with the response (\rho < 0.005). This suggests that spatial coordinates, observation volumes, and percentile values can meaningfully predict the PM2.5 concentration.

**Interpretation**
Northern locations tend to show lower PM2.5, while eastern location slightly increase it. This agrees with knowledge that many industries are established on the East Coast

More observations slightly reduce the average, perhaps that the lack of monitoring stations are inaccurately reflecting the true distribution.

**Limitations**
We are assuming linearity, homoscedasticity, and no multicollinearity, which are unlikely to be true with the given data.

### Regression Results

```{r regression_results}
# Filter to only PM2.5 observations
pm25_data <- data %>%
  filter(`Parameter Name` == "PM2.5 - Local Conditions") %>%
  select(`Arithmetic Mean`, Latitude, Longitude, Year, 
         `Observation Count`, `1st Max Value`, 
         `90th Percentile`, `10th Percentile`) %>%
  na.omit()

# Fit multiple linear regression model
lm_model <- lm(`Arithmetic Mean` ~ Latitude + Longitude + Year +
                 `Observation Count` + `1st Max Value` +
                 `90th Percentile` + `10th Percentile`,
               data = pm25_data)

# Create a tidy summary of the regression model
regression_summary <- tidy(lm_model) %>%
  mutate(p.value = round(p.value, 4),
         estimate = round(estimate, 4),
         std.error = round(std.error, 4),
         statistic = round(statistic, 4))

# Display the results in a table
kbl(regression_summary, caption = "Multiple Linear Regression Coefficients") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Column {data-width=500}
-------------------------------------

### Actual vs Predicted PM2.5

```{r actual_vs_predicted, fig.width=10, fig.height=10}
# Add fitted values and residuals
pm25_data$predicted <- predict(lm_model)
pm25_data$residuals <- resid(lm_model)

# Plot: Actual vs Predicted
ggplot(pm25_data, aes(x = `Arithmetic Mean`, y = predicted)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Actual vs Predicted PM2.5", x = "Actual", y = "Predicted") +
  theme_minimal()
```

### Residuals vs Fitted

```{r residuals, fig.width=10, fig.height=10}
# Residuals vs Fitted Plot
ggplot(pm25_data, aes(x = predicted, y = residuals)) +
  geom_point(alpha = 0.3, color = "darkorange") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted", x = "Predicted", y = "Residuals") +
  theme_minimal()
```

Spatial Analysis {#spatial-analysis data-navmenu="MLR" data-icon="fa-map"}
===================================== 

Column {data-width=1000}
-------------------------------------

### Spatial Prediction of PM2.5 Concentrations

```{r spatial, fig.width=14, fig.height=12}
# Create color palette
pal <- colorNumeric(
  palette = "plasma",
  domain = pm25_data$predicted
)

# Create leaflet map
leaflet(pm25_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    color = ~pal(predicted),
    stroke = FALSE,
    fillOpacity = 0.8,
    label = ~paste0("Predicted PM2.5: ", round(predicted, 2), " µg/m³")
  ) %>%
  addLegend(
    "bottomright",
    pal = pal,
    values = ~predicted,
    title = "Predicted PM2.5 (µg/m³)",
    opacity = 1
  )
```

















Ridge Regression
===================================== 

Row {data-height=150}
-------------------------------------

### Compute Model {data-width=400}
```{r ridge_model}
# Load the packages
library(glmnet)
library(plotly)
library(reshape2)
library(lubridate)

# Dynamically detect column names for State Code, County Code, and Site Num
possible_state_cols <- c("State Code", "State_Code", "STATE_CODE")
possible_county_cols <- c("County Code", "County_Code", "COUNTY_CODE")
possible_site_cols <- c("Site Num", "Site_Num", "SITE_NUM")

state_col <- intersect(possible_state_cols, colnames(data))
county_col <- intersect(possible_county_cols, colnames(data))
site_col <- intersect(possible_site_cols, colnames(data))

if (length(state_col) == 0 || length(county_col) == 0 || length(site_col) == 0) {
  stop("Error: Could not find columns for State Code, County Code, or Site Num. Available columns: ", paste(colnames(data), collapse = ", "))
}

# Use the detected column names
state_col <- state_col[1]
county_col <- county_col[1]
site_col <- site_col[1]

# Filter to PM2.5 24-hour block average observations and select relevant predictors
ridge_data <- data %>%
  filter(`Parameter Name` == "PM2.5 - Local Conditions" & `Sample Duration` == "24-HR BLK AVG") %>%
  mutate(Month = lubridate::month(as.Date(`Date of Last Change`))) %>%
  select(`Arithmetic Mean`, `1st Max Value`, `90th Percentile`, `75th Percentile`, 
         `50th Percentile`, `Observation Count`, `Observation Percent`, 
         `Latitude`, `Longitude`, `Year`, `Month`, 
         !!sym(state_col), !!sym(county_col), !!sym(site_col))

# Aggregate meteorological data to avoid duplicate entries
meteo_data <- data %>%
  filter(`Parameter Name` %in% c("Outdoor Temperature", "Relative Humidity", 
                                 "Wind Speed - Resultant", "Barometric Pressure", 
                                 "Solar Radiation", "Rain/melt Precipitation")) %>%
  group_by(!!sym(state_col), !!sym(county_col), !!sym(site_col), `Parameter Name`) %>%
  summarise(`Arithmetic Mean` = mean(`Arithmetic Mean`, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = `Parameter Name`, values_from = `Arithmetic Mean`)

# Check number of sites before and after join
cat("**Number of Sites Before Join:** ", nrow(ridge_data), "  \n")
message()
ridge_data_before_join <- ridge_data
ridge_data <- ridge_data %>%
  left_join(meteo_data, by = c(state_col, county_col, site_col))

# Preserve Year and Month before na.omit()
ridge_data <- ridge_data %>%
  mutate(Year = ridge_data_before_join$Year,
         Month = ridge_data_before_join$Month)

ridge_data <- ridge_data %>% na.omit()
cat("**Number of Sites After Join:** ", nrow(ridge_data), "  \n")

# Check if ridge_data is empty
if (nrow(ridge_data) == 0) {
  message("Warning: No data available after joining meteorological variables. Using only PM2.5-related predictors.")
  ridge_data <- data %>%
    filter(`Parameter Name` == "PM2.5 - Local Conditions" & `Sample Duration` == "24-HR BLK AVG") %>%
    mutate(Month = lubridate::month(as.Date(`Date of Last Change`))) %>%
    select(`Arithmetic Mean`, `1st Max Value`, `90th Percentile`, `75th Percentile`, 
           `50th Percentile`, `Observation Count`, `Observation Percent`, 
           `Latitude`, `Longitude`, `Year`, `Month`, 
           !!sym(state_col), !!sym(county_col), !!sym(site_col)) %>%
    na.omit()
  cat("**Number of Sites (PM2.5 Fallback):** ", nrow(ridge_data), "  \n")
}

# Number of sites per state
sites_per_state <- ridge_data %>%
  group_by(!!sym(state_col)) %>%
  summarise(Count = n()) %>%
  mutate(State = as.character(!!sym(state_col)))
cat("**Sites per State:**  \n")
for (i in 1:nrow(sites_per_state)) {
  cat("State ", sites_per_state$State[i], ": ", sites_per_state$Count[i], " sites  \n")
}

# Dynamically select predictor columns that exist in ridge_data
predictor_cols <- intersect(
  c("Outdoor Temperature", "Relative Humidity", "Wind Speed - Resultant", 
    "Barometric Pressure", "Solar Radiation", "Rain/melt Precipitation", 
    "1st Max Value", "90th Percentile", "75th Percentile", "50th Percentile", 
    "Observation Count", "Observation Percent", "Latitude", "Longitude", "Year", "Month"),
  colnames(ridge_data)
)

# Ensure all predictor columns are numeric
ridge_data <- ridge_data %>%
  mutate(across(.cols = all_of(predictor_cols), .fns = as.numeric))

# Prepare the predictor matrix and response variable
X <- model.matrix(`Arithmetic Mean` ~ . - 1, data = ridge_data) # Exclude intercept
y <- ridge_data$`Arithmetic Mean`

# Perform cross-validation to select the optimal lambda
set.seed(123) # For reproducibility
cv_model <- cv.glmnet(X, y, alpha = 0) # alpha = 0 for Ridge

# Fit the final Ridge model with the optimal lambda
ridge_model <- glmnet(X, y, alpha = 0, lambda = cv_model$lambda.min)

# Extract coefficients
ridge_coefs <- coef(ridge_model) %>%
  as.matrix() %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Predictor") %>%
  rename(Coefficient = s0) %>%
  mutate(Coefficient = round(Coefficient, 4))

# Compute predictions for summary
ridge_data$predicted <- predict(ridge_model, newx = X)

# Display computation summary
cat("**Model Computed Successfully**  \n")
cat("**Predictors Used:**  \n")
# Split predictors into smaller chunks to avoid overflow
predictor_chunks <- split(predictor_cols, ceiling(seq_along(predictor_cols)/5))
for (chunk in predictor_chunks) {
  cat(paste(chunk, collapse = ", "), "  \n")
}
cat("**Selected Lambda (Cross-Validation):** ", round(cv_model$lambda.min, 4), "  \n")
cat("**Predicted PM2.5 Range:** ", round(min(ridge_data$predicted, na.rm = TRUE), 2), " to ", 
    round(max(ridge_data$predicted, na.rm = TRUE), 2), " µg/m³  \n")
```

### Average Predicted PM2.5 {data-width=200}
```{r value_box_avg_predicted}
# Compute average predicted PM2.5
avg_predicted <- mean(ridge_data$predicted, na.rm = TRUE)
valueBox(round(avg_predicted, 2), icon = "fa-poll", color = "info")
```

### Number of Sites Analyzed {data-width=200}
```{r value_box_num_sites}
# Compute number of unique sites
num_sites <- nrow(ridge_data)
valueBox(num_sites, icon = "fa-map-marker-alt", color = "success")
```

### Data Coverage Period {data-width=200}
```{r value_box_year}
# Show the year of data
year_range <- unique(ridge_data$Year)
valueBox(year_range, icon = "fa-calendar", color = "warning")
```

Row {data-height=600}
-------------------------------------
### 3D Visualization of PM2.5 Trends {data-width=800}

```{r interactive_3d_plot, fig.width=12, fig.height=8}
# Debug: Verify data availability
message("Columns in ridge_data for 3D plot: ", paste(colnames(ridge_data), collapse = ", "))
message("Unique States in ridge_data: ", paste(unique(ridge_data[[state_col]]), collapse = ", "))

# Prepare plot data
plot_data <- ridge_data %>%
  select(`Outdoor Temperature`, `Relative Humidity`, predicted, Latitude, `Wind Speed - Resultant`, 
         !!sym(state_col), !!sym(county_col), !!sym(site_col)) %>%
  mutate(across(.cols = everything(), .fns = as.numeric)) %>%
  na.omit()

message("Rows in plot_data for 3D plot: ", nrow(plot_data))

if (nrow(plot_data) == 0) {
  message("No data available for 3D plot with meteorological variables. Switching to PM2.5 predictors.")
  plot_data <- ridge_data %>%
    select(`1st Max Value`, `90th Percentile`, predicted, Latitude, `Observation Count`, 
           !!sym(state_col), !!sym(county_col), !!sym(site_col)) %>%
    mutate(across(.cols = everything(), .fns = as.numeric)) %>%
    na.omit()
  message("Rows in plot_data (PM2.5 fallback): ", nrow(plot_data))
  
  if (nrow(plot_data) == 0) {
    stop("Error: No data available for 3D plot even with PM2.5 predictors.")
  }

  # Get unique states
  states <- unique(plot_data[[state_col]])
  message("Number of states: ", length(states))

  if (length(states) == 0) {
    stop("Error: No states available for dropdown menu.")
  }

  # Create a trace for each state, default to all visible
  traces <- lapply(states, function(state) {
    state_data <- plot_data[plot_data[[state_col]] == state, ]
    list(
      x = state_data$`1st Max Value`,
      y = state_data$`90th Percentile`,
      z = state_data$predicted,
      type = "scatter3d",
      mode = "markers",
      marker = list(
        size = 5,
        opacity = 0.7,
        color = state_data$Latitude,
        colorscale = "Viridis"
      ),
      text = paste("Site: ", state_data[[state_col]], "-", state_data[[county_col]], "-", state_data[[site_col]], 
                   "<br>PM2.5: ", round(state_data$predicted, 2)),
      hoverinfo = "text",
      name = as.character(state),
      visible = TRUE # Set all traces to visible by default
    )
  })

  # Create dropdown buttons
  buttons <- list()
  buttons[[1]] <- list(
    method = "restyle",
    args = list("visible", rep(TRUE, length(states))),
    label = "All States"
  )
  for (i in seq_along(states)) {
    vis <- rep(FALSE, length(states))
    vis[i] <- TRUE
    buttons[[i + 1]] <- list(
      method = "restyle",
      args = list("visible", vis),
      label = as.character(states[i])
    )
  }

  # Debug: Check the structure of buttons
  message("Number of buttons created: ", length(buttons))

  # Build the plot
  plot_3d <- plot_ly() %>%
    htmlwidgets::onRender("function(el) { el.on('plotly_restyle', function() { el.layout.updatemenus[0].active = -1; }); }")

  # Add traces dynamically
  for (i in seq_along(traces)) {
    plot_3d <- plot_3d %>% add_trace(
      x = traces[[i]]$x,
      y = traces[[i]]$y,
      z = traces[[i]]$z,
      type = traces[[i]]$type,
      mode = traces[[i]]$mode,
      marker = traces[[i]]$marker,
      text = traces[[i]]$text,
      hoverinfo = traces[[i]]$hoverinfo,
      name = traces[[i]]$name,
      visible = traces[[i]]$visible
    )
  }

  plot_3d <- plot_3d %>%
    layout(
      title = list(text = "Predicted PM2.5 by Max Value, 90th Percentile, and Latitude", pad = list(t = 40)),
      scene = list(
        xaxis = list(title = "1st Max Value (µg/m³)"),
        yaxis = list(title = "90th Percentile (µg/m³)"),
        zaxis = list(title = "Predicted PM2.5 (µg/m³)"),
        camera = list(eye = list(x = 1.5, y = 1.5, z = 1))
      ),
      showlegend = TRUE,
      updatemenus = list(
        list(
          y = 0.9,
          buttons = buttons,
          active = 0 # Set "All States" as the default active button
        )
      )
    )
} else {
  # Get unique states
  states <- unique(plot_data[[state_col]])
  message("Number of states: ", length(states))

  if (length(states) == 0) {
    stop("Error: No states available for dropdown menu.")
  }

  # Create a trace for each state, default to all visible
  traces <- lapply(states, function(state) {
    state_data <- plot_data[plot_data[[state_col]] == state, ]
    list(
      x = state_data$`Outdoor Temperature`,
      y = state_data$`Relative Humidity`,
      z = state_data$predicted,
      type = "scatter3d",
      mode = "markers",
      marker = list(
        size = 5,
        opacity = 0.7,
        color = state_data$Latitude,
        colorscale = "Viridis"
      ),
      text = paste("Site: ", state_data[[state_col]], "-", state_data[[county_col]], "-", state_data[[site_col]], 
                   "<br>PM2.5: ", round(state_data$predicted, 2)),
      hoverinfo = "text",
      name = as.character(state),
      visible = TRUE # Set all traces to visible by default
    )
  })

  # Create dropdown buttons
  buttons <- list()
  buttons[[1]] <- list(
    method = "restyle",
    args = list("visible", rep(TRUE, length(states))),
    label = "All States"
  )
  for (i in seq_along(states)) {
    vis <- rep(FALSE, length(states))
    vis[i] <- TRUE
    buttons[[i + 1]] <- list(
      method = "restyle",
      args = list("visible", vis),
      label = as.character(states[i])
    )
  }

  # Debug: Check the structure of buttons
  message("Number of buttons created: ", length(buttons))

  # Build the plot
  plot_3d <- plot_ly() %>%
    htmlwidgets::onRender("function(el) { el.on('plotly_restyle', function() { el.layout.updatemenus[0].active = -1; }); }")

  # Add traces dynamically
  for (i in seq_along(traces)) {
    plot_3d <- plot_3d %>% add_trace(
      x = traces[[i]]$x,
      y = traces[[i]]$y,
      z = traces[[i]]$z,
      type = traces[[i]]$type,
      mode = traces[[i]]$mode,
      marker = traces[[i]]$marker,
      text = traces[[i]]$text,
      hoverinfo = traces[[i]]$hoverinfo,
      name = traces[[i]]$name,
      visible = traces[[i]]$visible
    )
  }

  plot_3d <- plot_3d %>%
    layout(
      title = list(text = "Predicted PM2.5 by Temperature, Humidity, and Latitude", pad = list(t = 40)),
      scene = list(
        xaxis = list(title = "Outdoor Temperature (°F)"),
        yaxis = list(title = "Relative Humidity (%)"),
        zaxis = list(title = "Predicted PM2.5 (µg/m³)"),
        camera = list(eye = list(x = 1.5, y = 1.5, z = 1))
      ),
      showlegend = TRUE,
      updatemenus = list(
        list(
          y = 0.9,
          buttons = buttons,
          active = 0 # Set "All States" as the default active button
        )
      )
    )
}

plot_3d
```




### Model Summary {data-width=400}

**Research Question**: Can Ridge Regression effectively predict PM2.5 concentrations by accounting for multicollinearity among meteorological factors (e.g., temperature, relative humidity, wind speed) and air quality metrics, and how do these factors interact spatially and temporally to influence PM2.5 levels across monitoring sites?

**Modeling Approach**:  
- **Response Variable**: Arithmetic Mean (PM2.5, 24-hour block average)  
- **Predictor Variables**: Outdoor Temperature, Relative Humidity, Wind Speed - Resultant, Barometric Pressure, Solar Radiation, Rain/melt Precipitation, 1st Max Value, 90th Percentile, 75th Percentile, 50th Percentile, Observation Count, Observation Percent, Latitude, Longitude, Year, Month (derived from Date of Last Change)  

- **Justification**: Meteorological factors influence PM2.5 through atmospheric chemistry and dispersion, while air quality metrics capture exposure levels. Spatial and temporal variables account for geographic and seasonal trends. Ridge Regression’s L2 regularization addresses multicollinearity among correlated predictors (e.g., temperature and humidity).

**Discussion**: Our Ridge Regression model examines whether PM2.5 (24 hour block average, *Arithmetic Mean*) can be predicted by spatial data, weather conditions, and air quality. This addresses the issues of multicollinearity from the MLR analysis.

**Interpretation**: Meteorological conditions and air quality metrics jointly determine the PM2.5 levels, but site-level, country-level, and state-level factors (linked to local emission sources and regulations) also influence the PM2.5 levels. This regularization ensured temperature, humidity, and percentiles do not dominate the model due to multicolilinearity.

#### Model Coefficients
```{r ridge_coefficients}
# Display the coefficients in a table
kbl(ridge_coefs, caption = "Ridge Regression Coefficients (lambda = min)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Row {data-height=400}
-------------------------------------

### Actual vs Predicted PM2.5 (Ridge) {data-width=500}
```{r ridge_actual_vs_predicted, fig.width=8, fig.height=8}
# Plot: Actual vs Predicted
ggplot(ridge_data, aes(x = `Arithmetic Mean`, y = predicted)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Actual vs Predicted PM2.5 (Ridge)", x = "Actual PM2.5 (µg/m³)", y = "Predicted PM2.5 (µg/m³)") +
  theme_minimal()
```

### Predictor Correlation Heatmap {data-width=500}
```{r ridge_correlation_heatmap, fig.width=10, fig.height=10}
# Select available columns for correlation matrix
available_cols <- intersect(
  c("Outdoor Temperature", "Relative Humidity", "Wind Speed - Resultant", 
    "Barometric Pressure", "Solar Radiation", "Rain/melt Precipitation", 
    "1st Max Value", "90th Percentile", "75th Percentile", "50th Percentile", 
    "Observation Count", "Observation Percent", "Latitude", "Longitude", "Year", "Month"),
  colnames(ridge_data)
)

# Calculate correlation matrix for available predictors
cor_matrix <- cor(ridge_data[, available_cols])

# Create a heatmap with a cool color scheme
ggplot(data = reshape2::melt(cor_matrix), aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "Correlation Heatmap of Predictors", x = "", y = "")
```






LOESS Fit 
===================================== 
```{r loess_setup, include=FALSE}
library(highcharter)
# Define custom Highcharter theme
thm <- hc_theme(
  colors = c("#1a6ecc", "#434348", "#90ed7d"),
  chart = list(
    backgroundColor = "transparent",
    style = list(fontFamily = "Source Sans Pro")
  ),
  xAxis = list(gridLineWidth = 1)
)
```

Column {data-width=600}
-------------------------------------

### LOESS Fit for Seasonal PM2.5 Trends

**Research Question**: How do seasonal variations in PM2.5 concentrations manifest at a specific monitoring site in 2024?

**Modeling Approach**:  
- **Response Variable**: Arithmetic Mean (PM2.5 concentration)  
- **Predictor Variable**: Month (derived from 1st Max DateTime)  
- **Method**: LOESS smoothing with span = 0.75.

**Justification**: LOESS captures non-linear seasonal patterns in air quality data.

**Variables in Final Model**:  
- **Arithmetic Mean**: PM2.5 concentration.  
- **Month**: Numeric value (1-12) from `1st Max DateTime`.

**Discussion**: Analyzing how seasonal patterns affect PM2.t concentrations at monitoring site in 2024, we see that PM2.5 concentrations fluctuate across months, reflecting known seasonal factors influencing pollutants. Since LOESS adapts to non-linear seasonal changes, we can see a clearer picture of when air quality deteriorates or improves throughout the year.

### LOESS Model Visualization

```{r loess_fit, fig.height=8}
loess_data <- data %>%
  filter(`Parameter Name` == "PM2.5 - Local Conditions", `Site Num` == data$`Site Num`[1]) %>%
  mutate(Month = as.numeric(format(as.Date(`1st Max DateTime`), "%m"))) %>%
  select(`Arithmetic Mean`, Month) %>%
  na.omit()

loess_model <- loess(`Arithmetic Mean` ~ Month, data = loess_data, span = 0.75)
loess_data$predicted <- predict(loess_model)

highchart() %>%
  hc_add_series(name = "Observed", type = "scatter", 
                data = list_parse(data.frame(x = loess_data$Month, y = loess_data$`Arithmetic Mean`)),
                color = "#90ed7d", marker = list(radius = 4)) %>%
  hc_add_series(name = "LOESS Fit", type = "line", 
                data = list_parse(data.frame(x = loess_data$Month, y = loess_data$predicted)),
                color = "#1a6ecc", lineWidth = 2) %>%
  hc_xAxis(title = list(text = "Month"), categories = month.abb, gridLineWidth = 1) %>%
  hc_yAxis(title = list(text = "PM2.5 Concentration (µg/m³)")) %>%
  hc_title(text = "Seasonal PM2.5 Trends with LOESS Fit") %>%
  hc_add_theme(thm)
```

Column {.tabset data-width=400}
-------------------------------------
### Summary Statistics
```{r}
summary_stats <- loess_data %>%
  summarise(
    Mean_PM2.5 = round(mean(`Arithmetic Mean`), 2),
    SD_PM2.5 = round(sd(`Arithmetic Mean`), 2),
    Min_PM2.5 = round(min(`Arithmetic Mean`), 2),
    Max_PM2.5 = round(max(`Arithmetic Mean`), 2)
  ) %>%
  tidyr::pivot_longer(everything(), names_to = "Metric", values_to = "Value")
kbl(summary_stats, caption = "Summary Statistics for PM2.5 at Selected Site") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```
### Distribution by Month
```{r}
highchart() %>%
  hc_add_series(name = "PM2.5", type = "scatter",
                data = list_parse(data.frame(x = loess_data$Month, y = loess_data$`Arithmetic Mean`)),
                color = "#434348", marker = list(radius = 4)) %>%
  hc_xAxis(title = list(text = "Month"), categories = month.abb) %>%
  hc_yAxis(title = list(text = "PM2.5 Concentration (µg/m³)")) %>%
  hc_title(text = "PM2.5 Distribution by Month") %>%
  hc_add_theme(thm)
```







kNN Classification 
===================================== 
```{r knn_setup, include=FALSE}
library(caret)
```

Column {data-width=600}
-------------------------------------

### k-Nearest Neighbors Classification

**Research Question**: Can we classify monitoring sites as urban or rural based on air quality and spatial metrics in 2024?

**Modeling Approach**:  
- **Response Variable**: Site Type (Urban if in a Core-Based Statistical Area, else Rural)  
- **Predictor Variables**: Arithmetic Mean, 1st Max Value, 90th Percentile, Observation Count, Latitude, Longitude  
- **Method**: kNN classification with k optimized via cross-validation.

**Justification**: kNN classifies sites by leveraging similarities in air quality and spatial features, which differ between urban and rural areas.

**Variables in Final Model**:  
- **Site Type**: Binary response (Urban/Rural).  
- **Arithmetic Mean**, **1st Max Value**, **90th Percentile**, **Observation Count**, **Latitude**, **Longitude**: Predictors.

**Discussion**: The kNN model demonstrates that urban vs. rural site classification can be effectively performed using minimal air quality and spatial feature aching a 99.01% accuracy. This suggest that urban and rural sites are not overlapping regions in the feature space of PM2.5. We can determine the underlying pattern that urban areas have consistently higher levels of pollution and observation counts than rural areas, potentially due to high traffic and industrial activity, but also increase monitoring infrastructure.
### Model Performance

```{r knn_model}
knn_data <- data %>%
  filter(`Parameter Name` == "PM2.5 - Local Conditions") %>%
  mutate(Site_Type = factor(ifelse(is.na(`CBSA Name`), "Rural", "Urban"))) %>%
  select(Site_Type, `Arithmetic Mean`, `1st Max Value`, `90th Percentile`, 
         `Observation Count`, `Latitude`, `Longitude`) %>%
  na.omit()
set.seed(123)
train_idx <- sample(1:nrow(knn_data), 0.8 * nrow(knn_data))
train_data <- knn_data[train_idx, ]
test_data <- knn_data[-train_idx, ]
train_control <- trainControl(method = "cv", number = 5)
knn_model <- train(Site_Type ~ ., data = train_data, method = "knn", 
                   trControl = train_control, tuneGrid = expand.grid(k = 1:10))
knn_pred <- predict(knn_model, test_data)
results <- data.frame(Actual = test_data$Site_Type, Predicted = knn_pred)
accuracy <- mean(results$Actual == results$Predicted)
summary_table <- data.frame(
  Metric = c("Total Test Samples", "Overall Accuracy", "Optimal k"),
  Value = c(nrow(results), paste0(round(accuracy * 100, 2), "%"), knn_model$bestTune$k)
)
kbl(summary_table, caption = "kNN Model Performance") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```
Column {.tabset data-width=400}
-------------------------------------
### Confusion Matrix
```{r}
library(highcharter)
conf_matrix <- as.data.frame(table(Actual = results$Actual, Predicted = results$Predicted))
# Prepare data for heatmap with tooltips
heatmap_data <- lapply(1:nrow(conf_matrix), function(i) {
  list(
    x = as.numeric(factor(conf_matrix$Actual[i], levels = levels(conf_matrix$Actual))) - 1,
    y = as.numeric(factor(conf_matrix$Predicted[i], levels = levels(conf_matrix$Predicted))) - 1,
    value = conf_matrix$Freq[i],
    custom = list(actual = conf_matrix$Actual[i], predicted = conf_matrix$Predicted[i])
  )
})
# Define a fancier color gradient
color_stops <- list(
  list(0, "#ffffff"),
  list(0.5, "#ff9999"),
  list(1, "#d01010")
)
highchart() %>%
  hc_chart(type = "heatmap", backgroundColor = "transparent") %>%
  hc_add_series(name = "Confusion Matrix", type = "heatmap", data = heatmap_data, 
                borderColor = "#434348", borderWidth = 1,
                dataLabels = list(enabled = TRUE, color = "#ffffff", 
                                  style = list(fontFamily = "Open Sans", fontSize = "14px", 
                                               textOutline = "1px contrast"))) %>%
  hc_xAxis(title = list(text = "Actual", style = list(fontFamily = "Open Sans")), 
           categories = levels(conf_matrix$Actual), 
           gridLineWidth = 1, 
           labels = list(style = list(fontFamily = "Open Sans"))) %>%
  hc_yAxis(title = list(text = "Predicted", style = list(fontFamily = "Open Sans")), 
           categories = levels(conf_matrix$Predicted), 
           labels = list(style = list(fontFamily = "Open Sans"))) %>%
  hc_colorAxis(stops = color_stops, min = 0, 
               labels = list(style = list(fontFamily = "Open Sans"))) %>%
  hc_title(text = "Confusion Matrix", style = list(fontFamily = "Open Sans", fontSize = "18px")) %>%
  hc_tooltip(formatter = JS("function() {
    return '<b>Actual:</b> ' + this.point.custom.actual + '<br>' +
           '<b>Predicted:</b> ' + this.point.custom.predicted + '<br>' +
           '<b>Count:</b> ' + this.point.value;
  }")) %>%
  hc_plotOptions(series = list(
    dataLabels = list(enabled = TRUE),
    states = list(hover = list(enabled = TRUE, brightness = 0.1)),
    shadow = list(color = "rgba(0,0,0,0.3)", width = 5, opacity = 0.2)
  )) %>%
  hc_add_theme(hc_theme_smpl())
```
### Distribution Plot
```{r}
library(highcharter)
# Prepare data for Highcharter: group by Actual and Predicted to get counts
dist_data <- results %>%
  group_by(Actual, Predicted) %>%
  summarise(Count = n()) %>%
  ungroup()

# Create a list of series for Highcharter
series_data <- lapply(unique(dist_data$Predicted), function(pred) {
  subset <- dist_data %>% filter(Predicted == pred)
  list(
    name = pred,
    data = lapply(1:nrow(subset), function(i) {
      list(
        x = as.numeric(factor(subset$Actual[i], levels = levels(dist_data$Actual))) - 1,
        y = subset$Count[i],
        custom = list(actual = subset$Actual[i], predicted = subset$Predicted[i])
      )
    }),
    color = ifelse(pred == "Urban", "#d01010", "#90ed7d")
  )
})

highchart() %>%
  hc_chart(type = "column") %>%
  hc_add_series_list(series_data) %>%
  hc_xAxis(title = list(text = "Actual Site Type", style = list(fontFamily = "Open Sans")), 
           categories = levels(dist_data$Actual), 
           labels = list(style = list(fontFamily = "Open Sans"))) %>%
  hc_yAxis(title = list(text = "Count", style = list(fontFamily = "Open Sans")), 
           labels = list(style = list(fontFamily = "Open Sans"))) %>%
  hc_title(text = "Classification Distribution", style = list(fontFamily = "Open Sans", fontSize = "16px")) %>%
  hc_plotOptions(column = list(
    grouping = TRUE, 
    pointPadding = 0.1, 
    groupPadding = 0.2,
    dataLabels = list(enabled = TRUE, 
                      style = list(fontFamily = "Open Sans", fontSize = "12px", 
                                   textOutline = "1px contrast"),
                      color = "#ffffff"),
    states = list(hover = list(enabled = TRUE, brightness = 0.1))
  )) %>%
  hc_tooltip(formatter = JS("function() {
    return '<b>Actual:</b> ' + this.point.custom.actual + '<br>' +
           '<b>Predicted:</b> ' + this.point.custom.predicted + '<br>' +
           '<b>Count:</b> ' + this.y;
  }")) %>%
  hc_add_theme(hc_theme_smpl())
```











Naive Bayes
===================================== 

## Overview

- **Research Question:** Can we predict the pollutant type (`Parameter Name`) based on statistical features?
- **Modeling Approach:** Naive Bayes classifier, using predictors: Arithmetic Mean, 1st Max Value, 90th Percentile, Observation Count, etc.
- **Justification:** Naive Bayes is effective for multi-class classification with categorical outcomes.

```{r}
library(e1071) # For naive bayes

# Selecting relvant columns and removing NANs
nb_data <- data %>%
  select(`Parameter Name`,
         `Arithmetic Mean`, `1st Max Value`, `90th Percentile`, `Observation Count`,
         `POC`, `Latitude`, `Longitude`,
         `Sample Duration`, `Metric Used`, `Method Name`, `Observation Percent`,
         `Completeness Indicator`, `Required Day Count`, `Year`, `Units of Measure`, `Event Type`) %>%
  na.omit()

# Convert Response to factor
nb_data$`Parameter Name` = as.factor(nb_data$`Parameter Name`)

# Train/Test split
set.seed(1)
split_idx = sample(1:nrow(nb_data), size = 0.8 * nrow(nb_data))
train_data = nb_data[split_idx, ]
test_data = nb_data[-split_idx, ]

# Naive Bayes Model
nb_model = naiveBayes(`Parameter Name` ~ ., data = train_data)

# Predict on test set
nb_pred = predict(nb_model, test_data)

# Create summary table
results <- data.frame(
  Actual = test_data$`Parameter Name`,
  Predicted = nb_pred
)

accuracy <- mean(results$Actual == results$Predicted)
summary_table <- data.frame(
  Metric = c("Total Test Samples", "Overall Accuracy"),
  Value = c(nrow(results), paste0(round(accuracy * 100, 2), "%"))
)

# kbl(summary_table) %>%
#   kable_styling(bootstrap_options = c("striped", "hover"))
```

### Top 10 Most Frequently Predicted Pollutant Types

```{r}
predicted_summary <- results %>%
  count(Predicted, sort = TRUE) %>%
  head(10)

kbl(predicted_summary) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Misclassification Table (Top Errors)

```{r}
error_counts <- results %>%
  filter(Actual != Predicted) %>%
  count(Actual, Predicted, sort = TRUE) %>%
  top_n(5, n)

if (nrow(error_counts) > 0) {
  kbl(error_counts, caption = "Top 5 Misclassifications") %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
} else {
  print("No major misclassifications to display.")
}
```

Column {data-width=350}
-----------------------------------------------------------------------

```{r, height=300}
# plot

error_counts <- results %>%
  filter(Actual != Predicted) %>%
  count(Actual, Predicted, sort = TRUE) %>%
  top_n(10, n)

if (nrow(error_counts) > 0) {
  ggplot(error_counts, aes(x = reorder(paste(Actual, "->", Predicted), n), y = n, fill = n)) +
    geom_col() +
    coord_flip() +
    scale_fill_gradient(low = "#DDEBF7", high = "#2E75B6") +
    labs(title = "Top 10 Misclassifications", x = "Actual -> Predicted", y = "Count") +
    theme_minimal()
} else {
  print("No major misclassifications to display.")
}
```

### Discussion and Visualization

The model only achieved **`r round(accuracy * 100, 2)`%** accuracy when classifying monitoring sites as urban or rural on air quality and spatial features, indicating its assumption of feature independence does not hold. The model struggled to capture the complex patterns that separate urban and rural sites. Naive Bayes oversimplifies the relationships among environmental variables and we should use kNN instead.


------------------------------------------------------------------------

Logistic Regression
===================================== 

## Overview

- **Research Question:** What factors influence the likelihood of exceeding primary pollutant standards?
- **Modeling Approach:** 
  - **Response Variable:** Exceedance (binary: 1 if `Primary Exceedance Count` \> 0, else 0) 
  - **Predictor Variables:** Arithmetic Mean, 1st Max Value, Observation Count, Parameter Name (encoded)
- **Justification:** Logistic regression models the probability of a binary outcome based on predictor variables.

---

```{r}
logit_data = data %>%
  mutate(Exceedance = ifelse(`Primary Exceedance Count` > 0, 1, 0)) %>%
  select(Exceedance, `Arithmetic Mean`, `1st Max Value`, `Observation Count`, `Parameter Name`) %>%
  na.omit()

logit_data$Exceedance = as.factor(logit_data$Exceedance)
logit_data$`Parameter Name` = as.factor(logit_data$`Parameter Name`)


# test train split
set.seed(1)
split_idx = sample(1:nrow(logit_data), size = 0.8 * nrow(logit_data))
train_data = logit_data[split_idx, ]
test_data = logit_data[-split_idx, ]

# logistic regression
logit_model = glm(Exceedance ~ `Arithmetic Mean` + `1st Max Value` + `Observation Count` + `Parameter Name`,
                   data = train_data, family = binomial)
logit_probs = predict(logit_model, newdata = test_data, type = "response")
logit_pred = ifelse(logit_probs > 0.5, 1, 0)
actual = as.numeric(as.character(test_data$Exceedance))

# accuracy
accuracy = mean(logit_pred == actual)
summary_table = data.frame(
  Metric = c("Total Test Samples", "Overall Accuracy"),
  Value = c(length(actual), paste0(round(accuracy * 100, 2), "%"))
)

kbl(summary_table, caption = "Logistic Regression Model Performance") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

### Coefficient Summary

```{r}
tidy_summary = broom::tidy(logit_model)

kbl(tidy_summary, caption = "Logistic Regression Coefficients") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Discussion

The logistic regression model achieved an overall accuracy of **`r round(accuracy * 100, 2)`%** in predicting whether a site exceed primary pollutant standards. Observation Count showed significant positive effects on exceedance likelihood. However, most pollutant parameter categories (ozone, PM2.5, NO_2) had non-significant coefficients and were not meaningful to differentiate exceedance in this model. The results show that short-team peak pollution levels and observation intensity are the key features of exceedance risk.

---

## Visuals {data-width="350"}

### Coefficients

```{r}
library(broom)

tidy_summary = tidy_summary %>%
  mutate(
    conf.low = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error
  )

ggplot(tidy_summary, aes(x = estimate, y = reorder(term, estimate))) +
  geom_point(size = 2, color = "#2E75B6") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "#2E75B6") +
  labs(x = "Estimate (log-odds)", y = "Predictor") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  theme_minimal()
```

### Probalities vs Arithmetic Mean

```{r, message=FALSE, warning=FALSE}
plot_df = test_data
plot_df$Predicted_Prob = logit_probs

ggplot(plot_df, aes(x = `Arithmetic Mean`, y = Predicted_Prob)) +
  geom_point(aes(color = as.factor(Exceedance)), alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(x = "Arithmetic Mean Concentration",
       y = "Predicted Probability of Exceedance",
       color = "Actual Exceedance") +
  theme_minimal()
```