---
title: "U.S. Environment Insight"
author: "Yusi Yao, Xavier Akers, Keshavd"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    self_contained: true
    theme:
      version: 5
      bootswatch: darkly  
      base_font: -apple-system, BlinkMacSystemFont, "Segoe UI", Roboto, Helvetica, Arial, sans-serif  # GitHub's font stack
      code_font:
        google: JetBrains Mono
    css: custom.css
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(flexdashboard)
library(readr)
library(ggplot2)
library(leaflet)
library(dplyr)
library(kableExtra)
library(DT)
library(broom)
library(viridis)
# Install thematic and un-comment for themed static plots (i.e., ggplot2)
# thematic::thematic_rmd()
```

```{r data_load, echo=FALSE, message=FALSE}
data <- read_csv("data/annual_conc_by_monitor_2024.csv")
```







Introduction & Abstract
===================================== 

Column {data-width=350}
-------------------------------------

### Abstract

This dashboard analyzes the EPA Annual Concentration dataset for 2024, summarizing pollutant measurements across U.S. monitoring sites. We explore geographic patterns, pollutant trends, and apply statistical models to understand air quality outcomes and predictors.

### Data Dictionary

```{r data_dictionary}
variables <- colnames(data)

descriptions <- c(
  "State FIPS code", "County FIPS code", "Monitoring site number", "Parameter code for pollutant",
  "Point of contact", "Latitude coordinate", "Longitude coordinate", "Geodetic datum",
  "Pollutant name", "Sample duration", "Pollutant standard reference", "Metric used for analysis",
  "Measurement method", "Year of data", "Units of measure", "Event type recorded",
  "Number of valid observations", "Percent of valid observations", "Data completeness flag",
  "Number of valid days", "Number of required days", "Exceptional data count",
  "Null data count", "Primary standard exceedances", "Secondary standard exceedances",
  "Certification status", "Number of observations below detection limit",
  "Arithmetic mean concentration", "Arithmetic standard deviation", "Maximum observed value",
  "Datetime of max value", "Second max value", "Datetime of second max", "Third max value",
  "Datetime of third max", "Fourth max value", "Datetime of fourth max", 
  "First non-overlapping max", "Datetime of first non-overlapping max", 
  "Second non-overlapping max", "Datetime of second non-overlapping max", 
  "99th percentile", "98th percentile", "95th percentile", "90th percentile", 
  "75th percentile", "50th percentile", "10th percentile", "Local site name",
  "Monitoring site address", "State name", "County name", "City name",
  "Core-Based Statistical Area name", "Date of last data change"
)

data_dictionary <- data.frame(
  Variable = variables,
  Description = descriptions[1:length(variables)]
)

kbl(data_dictionary) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Column {data-width=650}
-------------------------------------

### Map of Monitoring Stations

```{r monitoring_map}
leaflet(data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~Longitude, lat = ~Latitude,
    radius = 1,
    popup = ~paste("Pollutant:", `Parameter Name`, "<br>Mean:", `Arithmetic Mean`)
  )
```







Multiple Regression {data-navmenu="MLR" data-icon="fa-chart-line"}
===================================== 

Column {data-width=650}
-------------------------------------

### Multiple Linear Regression

**Research Question**: Can we predict the annual average concentration (Arithmetic Mean) of PM2.5 based on location and temporal variables?

**Modeling Approach**:  
- **Response Variable**: Arithmetic Mean  
- **Predictor Variables**: Latitude, Longitude, Year, Observation Count, 1st Max Value, 90th Percentile, 10th Percentile  

**Justification**: This model aims to understand how geographical and temporal factors influence PM2.5 concentrations.

### Regression Results

```{r regression_results}
# Filter to only PM2.5 observations
pm25_data <- data %>%
  filter(`Parameter Name` == "PM2.5 - Local Conditions") %>%
  select(`Arithmetic Mean`, Latitude, Longitude, Year, 
         `Observation Count`, `1st Max Value`, 
         `90th Percentile`, `10th Percentile`) %>%
  na.omit()

# Fit multiple linear regression model
lm_model <- lm(`Arithmetic Mean` ~ Latitude + Longitude + Year +
                 `Observation Count` + `1st Max Value` +
                 `90th Percentile` + `10th Percentile`,
               data = pm25_data)

# Create a tidy summary of the regression model
regression_summary <- tidy(lm_model) %>%
  mutate(p.value = round(p.value, 4),
         estimate = round(estimate, 4),
         std.error = round(std.error, 4),
         statistic = round(statistic, 4))

# Display the results in a table
kbl(regression_summary, caption = "Multiple Linear Regression Coefficients") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Column {data-width=350}
-------------------------------------

### Actual vs Predicted PM2.5

```{r actual_vs_predicted, fig.width=8, fig.height=8}
# Add fitted values and residuals
pm25_data$predicted <- predict(lm_model)
pm25_data$residuals <- resid(lm_model)

# Plot: Actual vs Predicted
ggplot(pm25_data, aes(x = `Arithmetic Mean`, y = predicted)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Actual vs Predicted PM2.5", x = "Actual", y = "Predicted") +
  theme_minimal()
```

### Residuals vs Fitted

```{r residuals, fig.width=8, fig.height=8}
# Residuals vs Fitted Plot
ggplot(pm25_data, aes(x = predicted, y = residuals)) +
  geom_point(alpha = 0.3, color = "darkorange") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted", x = "Predicted", y = "Residuals") +
  theme_minimal()
```

Spatial Analysis {#spatial-analysis data-navmenu="MLR" data-icon="fa-map"}
===================================== 

Column {data-width=1000}
-------------------------------------

### Spatial Prediction of PM2.5 Concentrations

```{r spatial, fig.width=14, fig.height=12}
# Create color palette
pal <- colorNumeric(
  palette = "plasma",
  domain = pm25_data$predicted
)

# Create leaflet map
leaflet(pm25_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    color = ~pal(predicted),
    stroke = FALSE,
    fillOpacity = 0.8,
    label = ~paste0("Predicted PM2.5: ", round(predicted, 2), " µg/m³")
  ) %>%
  addLegend(
    "bottomright",
    pal = pal,
    values = ~predicted,
    title = "Predicted PM2.5 (µg/m³)",
    opacity = 1
  )
```









Ridge Regression
===================================== 

Row {data-height=150}
-------------------------------------

### Compute Model {data-width=400}
```{r ridge_model}
# Ensure all required packages are loaded
if (!requireNamespace("glmnet", quietly = TRUE)) {
  message("Installing glmnet...")
  install.packages("glmnet")
}
if (!requireNamespace("plotly", quietly = TRUE)) {
  message("Installing plotly...")
  install.packages("plotly")
}
if (!requireNamespace("reshape2", quietly = TRUE)) {
  message("Installing reshape2...")
  install.packages("reshape2")
}
if (!requireNamespace("lubridate", quietly = TRUE)) {
  message("Installing lubridate...")
  install.packages("lubridate")
}

# Load the packages
library(glmnet)
library(plotly)
library(reshape2)
library(lubridate)

# Confirm packages are loaded
message("Packages loaded: glmnet, plotly, reshape2, lubridate")

# Dynamically detect column names for State Code, County Code, and Site Num
possible_state_cols <- c("State Code", "State_Code", "STATE_CODE")
possible_county_cols <- c("County Code", "County_Code", "COUNTY_CODE")
possible_site_cols <- c("Site Num", "Site_Num", "SITE_NUM")

state_col <- intersect(possible_state_cols, colnames(data))
county_col <- intersect(possible_county_cols, colnames(data))
site_col <- intersect(possible_site_cols, colnames(data))

if (length(state_col) == 0 || length(county_col) == 0 || length(site_col) == 0) {
  stop("Error: Could not find columns for State Code, County Code, or Site Num. Available columns: ", paste(colnames(data), collapse = ", "))
}

# Use the detected column names
state_col <- state_col[1]
county_col <- county_col[1]
site_col <- site_col[1]

# Filter to PM2.5 24-hour block average observations and select relevant predictors
ridge_data <- data %>%
  filter(`Parameter Name` == "PM2.5 - Local Conditions" & `Sample Duration` == "24-HR BLK AVG") %>%
  mutate(Month = lubridate::month(as.Date(`Date of Last Change`))) %>%
  select(`Arithmetic Mean`, `1st Max Value`, `90th Percentile`, `75th Percentile`, 
         `50th Percentile`, `Observation Count`, `Observation Percent`, 
         `Latitude`, `Longitude`, `Year`, `Month`, 
         !!sym(state_col), !!sym(county_col), !!sym(site_col))

# Aggregate meteorological data to avoid duplicate entries
meteo_data <- data %>%
  filter(`Parameter Name` %in% c("Outdoor Temperature", "Relative Humidity", 
                                 "Wind Speed - Resultant", "Barometric Pressure", 
                                 "Solar Radiation", "Rain/melt Precipitation")) %>%
  group_by(!!sym(state_col), !!sym(county_col), !!sym(site_col), `Parameter Name`) %>%
  summarise(`Arithmetic Mean` = mean(`Arithmetic Mean`, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = `Parameter Name`, values_from = `Arithmetic Mean`)

# Check number of sites before and after join
cat("**Number of Sites Before Join:** ", nrow(ridge_data), "  \n")
ridge_data_before_join <- ridge_data
ridge_data <- ridge_data %>%
  left_join(meteo_data, by = c(state_col, county_col, site_col))

# Preserve Year and Month before na.omit()
ridge_data <- ridge_data %>%
  mutate(Year = ridge_data_before_join$Year,
         Month = ridge_data_before_join$Month)

ridge_data <- ridge_data %>% na.omit()
cat("**Number of Sites After Join:** ", nrow(ridge_data), "  \n")

# Check if ridge_data is empty
if (nrow(ridge_data) == 0) {
  message("Warning: No data available after joining meteorological variables. Using only PM2.5-related predictors.")
  ridge_data <- data %>%
    filter(`Parameter Name` == "PM2.5 - Local Conditions" & `Sample Duration` == "24-HR BLK AVG") %>%
    mutate(Month = lubridate::month(as.Date(`Date of Last Change`))) %>%
    select(`Arithmetic Mean`, `1st Max Value`, `90th Percentile`, `75th Percentile`, 
           `50th Percentile`, `Observation Count`, `Observation Percent`, 
           `Latitude`, `Longitude`, `Year`, `Month`, 
           !!sym(state_col), !!sym(county_col), !!sym(site_col)) %>%
    na.omit()
  cat("**Number of Sites (PM2.5 Fallback):** ", nrow(ridge_data), "  \n")
}

# Number of sites per state
sites_per_state <- ridge_data %>%
  group_by(!!sym(state_col)) %>%
  summarise(Count = n()) %>%
  mutate(State = as.character(!!sym(state_col)))
cat("**Sites per State:**  \n")
for (i in 1:nrow(sites_per_state)) {
  cat("State ", sites_per_state$State[i], ": ", sites_per_state$Count[i], " sites  \n")
}

# Dynamically select predictor columns that exist in ridge_data
predictor_cols <- intersect(
  c("Outdoor Temperature", "Relative Humidity", "Wind Speed - Resultant", 
    "Barometric Pressure", "Solar Radiation", "Rain/melt Precipitation", 
    "1st Max Value", "90th Percentile", "75th Percentile", "50th Percentile", 
    "Observation Count", "Observation Percent", "Latitude", "Longitude", "Year", "Month"),
  colnames(ridge_data)
)

# Ensure all predictor columns are numeric
ridge_data <- ridge_data %>%
  mutate(across(.cols = all_of(predictor_cols), .fns = as.numeric))

# Prepare the predictor matrix and response variable
X <- model.matrix(`Arithmetic Mean` ~ . - 1, data = ridge_data) # Exclude intercept
y <- ridge_data$`Arithmetic Mean`

# Perform cross-validation to select the optimal lambda
set.seed(123) # For reproducibility
cv_model <- cv.glmnet(X, y, alpha = 0) # alpha = 0 for Ridge

# Fit the final Ridge model with the optimal lambda
ridge_model <- glmnet(X, y, alpha = 0, lambda = cv_model$lambda.min)

# Extract coefficients
ridge_coefs <- coef(ridge_model) %>%
  as.matrix() %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Predictor") %>%
  rename(Coefficient = s0) %>%
  mutate(Coefficient = round(Coefficient, 4))

# Compute predictions for summary
ridge_data$predicted <- predict(ridge_model, newx = X)

# Display computation summary
cat("**Model Computed Successfully**  \n")
cat("**Predictors Used:**  \n")
# Split predictors into smaller chunks to avoid overflow
predictor_chunks <- split(predictor_cols, ceiling(seq_along(predictor_cols)/5))
for (chunk in predictor_chunks) {
  cat(paste(chunk, collapse = ", "), "  \n")
}
cat("**Selected Lambda (Cross-Validation):** ", round(cv_model$lambda.min, 4), "  \n")
cat("**Predicted PM2.5 Range:** ", round(min(ridge_data$predicted, na.rm = TRUE), 2), " to ", 
    round(max(ridge_data$predicted, na.rm = TRUE), 2), " µg/m³  \n")
```

### Average Predicted PM2.5 {data-width=200}
```{r value_box_avg_predicted}
# Compute average predicted PM2.5
avg_predicted <- mean(ridge_data$predicted, na.rm = TRUE)
valueBox(round(avg_predicted, 2), icon = "fa-poll", color = "info")
```

### Number of Sites Analyzed {data-width=200}
```{r value_box_num_sites}
# Compute number of unique sites
num_sites <- nrow(ridge_data)
valueBox(num_sites, icon = "fa-map-marker-alt", color = "success")
```

### Data Coverage Period {data-width=200}
```{r value_box_year}
# Show the year of data
year_range <- unique(ridge_data$Year)
valueBox(year_range, icon = "fa-calendar", color = "warning")
```

Row {data-height=600}
-------------------------------------

### 3D Visualization of PM2.5 Trends {data-width=800}

```{r interactive_3d_plot, fig.width=12, fig.height=8}
# Debug: Verify data availability
message("Columns in ridge_data for 3D plot: ", paste(colnames(ridge_data), collapse = ", "))
message("Unique States in ridge_data: ", paste(unique(ridge_data[[state_col]]), collapse = ", "))

# Prepare plot data
plot_data <- ridge_data %>%
  select(`Outdoor Temperature`, `Relative Humidity`, predicted, Latitude, `Wind Speed - Resultant`, 
         !!sym(state_col), !!sym(county_col), !!sym(site_col)) %>%
  mutate(across(.cols = everything(), .fns = as.numeric)) %>%
  na.omit()

message("Rows in plot_data for 3D plot: ", nrow(plot_data))

if (nrow(plot_data) == 0) {
  message("No data available for 3D plot with meteorological variables. Switching to PM2.5 predictors.")
  plot_data <- ridge_data %>%
    select(`1st Max Value`, `90th Percentile`, predicted, Latitude, `Observation Count`, 
           !!sym(state_col), !!sym(county_col), !!sym(site_col)) %>%
    mutate(across(.cols = everything(), .fns = as.numeric)) %>%
    na.omit()
  message("Rows in plot_data (PM2.5 fallback): ", nrow(plot_data))
  
  if (nrow(plot_data) == 0) {
    stop("Error: No data available for 3D plot even with PM2.5 predictors.")
  }

  # Get unique states
  states <- unique(plot_data[[state_col]])
  message("Number of states: ", length(states))

  if (length(states) == 0) {
    stop("Error: No states available for dropdown menu.")
  }

  # Create a trace for each state
  traces <- lapply(states, function(state) {
    state_data <- plot_data[plot_data[[state_col]] == state, ]
    list(
      x = state_data$`1st Max Value`,
      y = state_data$`90th Percentile`,
      z = state_data$predicted,
      type = "scatter3d",
      mode = "markers",
      marker = list(
        size = 5,
        opacity = 0.7,
        color = state_data$Latitude,
        colorscale = "Viridis"
      ),
      text = paste("Site: ", state_data[[state_col]], "-", state_data[[county_col]], "-", state_data[[site_col]], 
                   "<br>PM2.5: ", round(state_data$predicted, 2)),
      hoverinfo = "text",
      name = as.character(state),
      visible = (state == states[1]) # Only the first state is visible initially
    )
  })

  # Create dropdown buttons
  buttons <- list()
  buttons[[1]] <- list(
    method = "restyle",
    args = list("visible", rep(TRUE, length(states))),
    label = "All States"
  )
  for (i in seq_along(states)) {
    vis <- rep(FALSE, length(states))
    vis[i] <- TRUE
    buttons[[i + 1]] <- list(
      method = "restyle",
      args = list("visible", vis),
      label = as.character(states[i])
    )
  }

  # Debug: Check the structure of buttons
  message("Number of buttons created: ", length(buttons))

  # Build the plot
  plot_3d <- plot_ly() %>%
    htmlwidgets::onRender("function(el) { el.on('plotly_restyle', function() { el.layout.updatemenus[0].active = -1; }); }")

  # Add traces dynamically
  for (i in seq_along(traces)) {
    plot_3d <- plot_3d %>% add_trace(
      x = traces[[i]]$x,
      y = traces[[i]]$y,
      z = traces[[i]]$z,
      type = traces[[i]]$type,
      mode = traces[[i]]$mode,
      marker = traces[[i]]$marker,
      text = traces[[i]]$text,
      hoverinfo = traces[[i]]$hoverinfo,
      name = traces[[i]]$name,
      visible = traces[[i]]$visible
    )
  }

  plot_3d <- plot_3d %>%
    layout(
      title = list(text = "Predicted PM2.5 by Max Value, 90th Percentile, and Latitude", pad = list(t = 40)),
      scene = list(
        xaxis = list(title = "1st Max Value (µg/m³)"),
        yaxis = list(title = "90th Percentile (µg/m³)"),
        zaxis = list(title = "Predicted PM2.5 (µg/m³)"),
        camera = list(eye = list(x = 1.5, y = 1.5, z = 1))
      ),
      showlegend = TRUE,
      updatemenus = list(
        list(
          y = 0.9,
          buttons = buttons
        )
      )
    )
} else {
  # Get unique states
  states <- unique(plot_data[[state_col]])
  message("Number of states: ", length(states))

  if (length(states) == 0) {
    stop("Error: No states available for dropdown menu.")
  }

  # Create a trace for each state
  traces <- lapply(states, function(state) {
    state_data <- plot_data[plot_data[[state_col]] == state, ]
    list(
      x = state_data$`Outdoor Temperature`,
      y = state_data$`Relative Humidity`,
      z = state_data$predicted,
      type = "scatter3d",
      mode = "markers",
      marker = list(
        size = 5,
        opacity = 0.7,
        color = state_data$Latitude,
        colorscale = "Viridis"
      ),
      text = paste("Site: ", state_data[[state_col]], "-", state_data[[county_col]], "-", state_data[[site_col]], 
                   "<br>PM2.5: ", round(state_data$predicted, 2)),
      hoverinfo = "text",
      name = as.character(state),
      visible = (state == states[1]) # Only the first state is visible initially
    )
  })

  # Create dropdown buttons
  buttons <- list()
  buttons[[1]] <- list(
    method = "restyle",
    args = list("visible", rep(TRUE, length(states))),
    label = "All States"
  )
  for (i in seq_along(states)) {
    vis <- rep(FALSE, length(states))
    vis[i] <- TRUE
    buttons[[i + 1]] <- list(
      method = "restyle",
      args = list("visible", vis),
      label = as.character(states[i])
    )
  }

  # Debug: Check the structure of buttons
  message("Number of buttons created: ", length(buttons))

  # Build the plot
  plot_3d <- plot_ly() %>%
    htmlwidgets::onRender("function(el) { el.on('plotly_restyle', function() { el.layout.updatemenus[0].active = -1; }); }")

  # Add traces dynamically
  for (i in seq_along(traces)) {
    plot_3d <- plot_3d %>% add_trace(
      x = traces[[i]]$x,
      y = traces[[i]]$y,
      z = traces[[i]]$z,
      type = traces[[i]]$type,
      mode = traces[[i]]$mode,
      marker = traces[[i]]$marker,
      text = traces[[i]]$text,
      hoverinfo = traces[[i]]$hoverinfo,
      name = traces[[i]]$name,
      visible = traces[[i]]$visible
    )
  }

  plot_3d <- plot_3d %>%
    layout(
      title = list(text = "Predicted PM2.5 by Temperature, Humidity, and Latitude", pad = list(t = 40)),
      scene = list(
        xaxis = list(title = "Outdoor Temperature (°F)"),
        yaxis = list(title = "Relative Humidity (%)"),
        zaxis = list(title = "Predicted PM2.5 (µg/m³)"),
        camera = list(eye = list(x = 1.5, y = 1.5, z = 1))
      ),
      showlegend = TRUE,
      updatemenus = list(
        list(
          y = 0.9,
          buttons = buttons
        )
      )
    )
}

plot_3d
```

### Model Summary {data-width=400}

**Research Question**: Can Ridge Regression effectively predict PM2.5 concentrations by accounting for multicollinearity among meteorological factors (e.g., temperature, relative humidity, wind speed) and air quality metrics, and how do these factors interact spatially and temporally to influence PM2.5 levels across monitoring sites?

**Modeling Approach**:  
- **Response Variable**: Arithmetic Mean (PM2.5, 24-hour block average)  
- **Predictor Variables**: Outdoor Temperature, Relative Humidity, Wind Speed - Resultant, Barometric Pressure, Solar Radiation, Rain/melt Precipitation, 1st Max Value, 90th Percentile, 75th Percentile, 50th Percentile, Observation Count, Observation Percent, Latitude, Longitude, Year, Month (derived from Date of Last Change)  
- **Justification**: Meteorological factors influence PM2.5 through atmospheric chemistry and dispersion, while air quality metrics capture exposure levels. Spatial and temporal variables account for geographic and seasonal trends. Ridge Regression’s L2 regularization addresses multicollinearity among correlated predictors (e.g., temperature and humidity).

#### Model Coefficients
```{r ridge_coefficients}
# Display the coefficients in a table
kbl(ridge_coefs, caption = "Ridge Regression Coefficients (lambda = min)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Row {data-height=400}
-------------------------------------

### Actual vs Predicted PM2.5 (Ridge) {data-width=500}
```{r ridge_actual_vs_predicted, fig.width=8, fig.height=8}
# Plot: Actual vs Predicted
ggplot(ridge_data, aes(x = `Arithmetic Mean`, y = predicted)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Actual vs Predicted PM2.5 (Ridge)", x = "Actual PM2.5 (µg/m³)", y = "Predicted PM2.5 (µg/m³)") +
  theme_minimal()
```

### Predictor Correlation Heatmap {data-width=500}
```{r ridge_correlation_heatmap, fig.width=10, fig.height=10}
# Select available columns for correlation matrix
available_cols <- intersect(
  c("Outdoor Temperature", "Relative Humidity", "Wind Speed - Resultant", 
    "Barometric Pressure", "Solar Radiation", "Rain/melt Precipitation", 
    "1st Max Value", "90th Percentile", "75th Percentile", "50th Percentile", 
    "Observation Count", "Observation Percent", "Latitude", "Longitude", "Year", "Month"),
  colnames(ridge_data)
)

# Calculate correlation matrix for available predictors
cor_matrix <- cor(ridge_data[, available_cols])

# Create a heatmap with a cool color scheme
ggplot(data = reshape2::melt(cor_matrix), aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "Correlation Heatmap of Predictors", x = "", y = "")
```












# Natural Spline or LOESS

## Column {data-width="650"}

3.  Natural Cubic Spline or LOESS Fit

Research Question: How do PM2.5 concentrations vary over time within a
specific location?

Modeling Approach: - Response Variable: Arithmetic Mean - Predictor
Variable: Date (converted from 1st Max DateTime) - Method: Apply a
natural cubic spline or LOESS smoothing to model temporal trends.

Justification: This approach captures non-linear temporal patterns in
pollutant concentrations.

------------------------------------------------------------------------

# kNN

## Column {data-width="650"}

4.  k-Nearest Neighbors (kNN) Classification

Research Question: Can we classify days as ‘High’ or ‘Low’ pollution
based on observed metrics?

Modeling Approach: • Response Variable: Pollution Level (categorical:
‘High’ if Arithmetic Mean \> threshold, else ‘Low’) • Predictor
Variables: 1st Max Value, 90th Percentile, Observation Count, Latitude,
Longitude

Justification: kNN can classify pollution levels based on similarity in
observed features.

------------------------------------------------------------------------

Naive Bayes
===================================== 

## Overview

- **Research Question:** Can we predict the pollutant type (`Parameter Name`) based on statistical features?
- **Modeling Approach:** Naive Bayes classifier, using predictors: Arithmetic Mean, 1st Max Value, 90th Percentile, Observation Count, etc.
- **Justification:** Naive Bayes is effective for multi-class classification with categorical outcomes.


```{r}
library(e1071) # For naive bayes

# Selecting relvant columns and removing NANs
nb_data <- data %>%
  select(`Parameter Name`,
         `Arithmetic Mean`, `1st Max Value`, `90th Percentile`, `Observation Count`,
         `POC`, `Latitude`, `Longitude`,
         `Sample Duration`, `Metric Used`, `Method Name`, `Observation Percent`,
         `Completeness Indicator`, `Required Day Count`, `Year`, `Units of Measure`, `Event Type`) %>%
  na.omit()

# Convert Response to factor
nb_data$`Parameter Name` = as.factor(nb_data$`Parameter Name`)

# Train/Test split
set.seed(1)
split_idx = sample(1:nrow(nb_data), size = 0.8 * nrow(nb_data))
train_data = nb_data[split_idx, ]
test_data = nb_data[-split_idx, ]

# Naive Bayes Model
nb_model = naiveBayes(`Parameter Name` ~ ., data = train_data)

# Predict on test set
nb_pred = predict(nb_model, test_data)

# Create summary table
results <- data.frame(
  Actual = test_data$`Parameter Name`,
  Predicted = nb_pred
)

accuracy <- mean(results$Actual == results$Predicted)
summary_table <- data.frame(
  Metric = c("Total Test Samples", "Overall Accuracy"),
  Value = c(nrow(results), paste0(round(accuracy * 100, 2), "%"))
)

# kbl(summary_table) %>%
#   kable_styling(bootstrap_options = c("striped", "hover"))
```

### Top 10 Most Frequently Predicted Pollutant Types

```{r}
predicted_summary <- results %>%
  count(Predicted, sort = TRUE) %>%
  head(10)

kbl(predicted_summary) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Misclassification Table (Top Errors)

```{r}
error_counts <- results %>%
  filter(Actual != Predicted) %>%
  count(Actual, Predicted, sort = TRUE) %>%
  top_n(5, n)

if (nrow(error_counts) > 0) {
  kbl(error_counts, caption = "Top 5 Misclassifications") %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
} else {
  print("No major misclassifications to display.")
}
```

Column {data-width=350}
-----------------------------------------------------------------------

```{r, height=300}
# plot

error_counts <- results %>%
  filter(Actual != Predicted) %>%
  count(Actual, Predicted, sort = TRUE) %>%
  top_n(10, n)

if (nrow(error_counts) > 0) {
  ggplot(error_counts, aes(x = reorder(paste(Actual, "->", Predicted), n), y = n, fill = n)) +
    geom_col() +
    coord_flip() +
    scale_fill_gradient(low = "#DDEBF7", high = "#2E75B6") +
    labs(title = "Top 10 Misclassifications", x = "Actual -> Predicted", y = "Count") +
    theme_minimal()
} else {
  print("No major misclassifications to display.")
}
```

### Discussion and Visualization

The Naive Bayes classifier achieved an overall accuracy of **`r round(accuracy * 100, 2)`%**.



------------------------------------------------------------------------

Logistic Regression
===================================== 

## Overview

- **Research Question:** What factors influence the likelihood of exceeding primary pollutant standards?
- **Modeling Approach:** 
  - **Response Variable:** Exceedance (binary: 1 if `Primary Exceedance Count` \> 0, else 0) 
  - **Predictor Variables:** Arithmetic Mean, 1st Max Value, Observation Count, Parameter Name (encoded)
- **Justification:** Logistic regression models the probability of a binary outcome based on predictor variables.

---

```{r}
logit_data = data %>%
  mutate(Exceedance = ifelse(`Primary Exceedance Count` > 0, 1, 0)) %>%
  select(Exceedance, `Arithmetic Mean`, `1st Max Value`, `Observation Count`, `Parameter Name`) %>%
  na.omit()

logit_data$Exceedance = as.factor(logit_data$Exceedance)
logit_data$`Parameter Name` = as.factor(logit_data$`Parameter Name`)


# test train split
set.seed(1)
split_idx = sample(1:nrow(logit_data), size = 0.8 * nrow(logit_data))
train_data = logit_data[split_idx, ]
test_data = logit_data[-split_idx, ]

# logistic regression
logit_model = glm(Exceedance ~ `Arithmetic Mean` + `1st Max Value` + `Observation Count` + `Parameter Name`,
                   data = train_data, family = binomial)
logit_probs = predict(logit_model, newdata = test_data, type = "response")
logit_pred = ifelse(logit_probs > 0.5, 1, 0)
actual = as.numeric(as.character(test_data$Exceedance))

# accuracy
accuracy = mean(logit_pred == actual)
summary_table = data.frame(
  Metric = c("Total Test Samples", "Overall Accuracy"),
  Value = c(length(actual), paste0(round(accuracy * 100, 2), "%"))
)

kbl(summary_table, caption = "Logistic Regression Model Performance") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

### Coefficient Summary

```{r}
tidy_summary = broom::tidy(logit_model)

kbl(tidy_summary, caption = "Logistic Regression Coefficients") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Discussion

The logistic regression model achieved an overall accuracy of **`r round(accuracy * 100, 2)`%**.

---

## Visuals {data-width="350"}

### Coefficients

```{r}
library(broom)

tidy_summary = tidy_summary %>%
  mutate(
    conf.low = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error
  )

ggplot(tidy_summary, aes(x = estimate, y = reorder(term, estimate))) +
  geom_point(size = 2, color = "#2E75B6") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "#2E75B6") +
  labs(x = "Estimate (log-odds)", y = "Predictor") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  theme_minimal()
```

### Probalities vs Arithmetic Mean

```{r, message=FALSE, warning=FALSE}
plot_df = test_data
plot_df$Predicted_Prob = logit_probs

ggplot(plot_df, aes(x = `Arithmetic Mean`, y = Predicted_Prob)) +
  geom_point(aes(color = as.factor(Exceedance)), alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(x = "Arithmetic Mean Concentration",
       y = "Predicted Probability of Exceedance",
       color = "Actual Exceedance") +
  theme_minimal()
```