---
title: "Final Template"
author: "Yusi Yao, Xavier, Keshavd"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    self_contained: true
---

```{r setup, include=FALSE}
library(flexdashboard)
# Install thematic and un-comment for themed static plots (i.e., ggplot2)
# thematic::thematic_rmd()

```

```{r, echo=FALSE, message=FALSE}
library(readr)
data <- read_csv("data/annual_conc_by_monitor_2024.csv")
```

# Multiple Regression

Column {data-width=650}
-------------------------------------

### Description

1. Multiple Linear Regression

Research Question:
Can we predict the annual average concentration (Arithmetic Mean) of PM2.5 based on location and temporal variables?

Modeling Approach:
	•	Response Variable: Arithmetic Mean
	•	Predictor Variables: Latitude, Longitude, Year, Observation Count, 1st Max Value, 90th Percentile, 10th Percentile

Justification:
This model aims to understand how geographical and temporal factors influence PM2.5 concentrations.

---








# Ridge Regression

Column {data-width=650}
-------------------------------------
2. Ridge Regression (using glmnet)

Research Question:
Does incorporating multiple correlated predictors improve the prediction of PM2.5 concentrations?

Modeling Approach:
- Response Variable: Arithmetic Mean
- Predictor Variables: 1st Max Value, 2nd Max Value, 3rd Max Value, 4th Max Value, 90th Percentile, 75th Percentile, 50th Percentile, 10th Percentile, Observation Count

Justification:
Ridge regression is suitable for handling multicollinearity among predictors, potentially enhancing model performance.

---

# Natural Spline or LOESS

Column {data-width=650}
-------------------------------------

3. Natural Cubic Spline or LOESS Fit

Research Question:
How do PM2.5 concentrations vary over time within a specific location?

Modeling Approach:
- Response Variable: Arithmetic Mean
- Predictor Variable: Date (converted from 1st Max DateTime)
- Method: Apply a natural cubic spline or LOESS smoothing to model temporal trends.

Justification:
This approach captures non-linear temporal patterns in pollutant concentrations.

---

# kNN

Column {data-width=650}
-------------------------------------

4. k-Nearest Neighbors (kNN) Classification

Research Question:
Can we classify days as ‘High’ or ‘Low’ pollution based on observed metrics?

Modeling Approach:
	•	Response Variable: Pollution Level (categorical: ‘High’ if Arithmetic Mean > threshold, else ‘Low’)
	•	Predictor Variables: 1st Max Value, 90th Percentile, Observation Count, Latitude, Longitude

Justification:
kNN can classify pollution levels based on similarity in observed features.

---

# Naive Bayes

Column {data-width=650}
-------------------------------------
5. Naive Bayes Classification

Research Question:
Can we predict the Parameter Name (type of pollutant) based on statistical features?

Modeling Approach:
	- Response Variable: Parameter Name
	- Predictor Variables: Arithmetic Mean, 1st Max Value, 90th Percentile, Observation Count

Justification:
Naive Bayes is effective for multi-class classification problems with categorical outcomes.

---

# Logistic Regression

Column {data-width=650}
-------------------------------------

6. Logistic Regression Classification

Research Question:
What factors influence the likelihood of exceeding primary pollutant standards?

Modeling Approach:
	- Response Variable: Exceedance (binary: 1 if Primary Exceedance Count > 0, else 0)
	- Predictor Variables: Arithmetic Mean, 1st Max Value, Observation Count, Parameter Name (encoded)

Justification:
Logistic regression models the probability of a binary outcome based on predictor variables.


Page 1 {data-navmenu="Menu A" data-navmenu-icon="fa-bookmark"}
=====================================


Page 2 {data-navmenu="Menu A"}
=====================================  


Page 3 {data-navmenu="Menu B" data-navmenu="fa-cog"}
=====================================


Page 4 {data-navmenu="Menu B"}
=====================================  

Page 1
===================================== 

You can link to a dashboard page with either of the following syntaxes:

[Page 2]

[Page Two](#page-2)

The second syntax is used when you want a custom name for the link 
(rather than just using the page title).

### Chart A
    
```{r}
```

Page 2
=====================================     

### Chart B
    
```{r}
```


